---
title: "Divvy Trip Data Analysis"
output: pdf_document
---

## Setup and Data Loading

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE)
```
```{r load-data}
divvy_data <- read.csv("Divvy_Trips_2019_Q1.csv", header = TRUE)
```

# Display the first 6 rows of the data frame (default behavior of head())
```{r display-data, echo=FALSE}
head(divvy_data)
```
# You can also specify the number of rows you want to see, e.g., the first 10 rows:
# print(head(divvy_data, n = 10))

*** Now I want to print the columns names of the data frame. ***

```{r column-names, echo=FALSE}
colnames(divvy_data)
```

*** Now I want to print the structure of the data frame. ***

```{r structure-data, echo=FALSE}
str(divvy_data)

```

*** Now I want to print the summary of the data frame. ***

```{r summary-data, echo=FALSE}
summary(divvy_data)
```

*** Now I need to check the number of usertypes in the data frame. ***

```{r unique-usertypes, echo=FALSE}
unique(divvy_data$usertype)
```
*** I need to see the number of unique Subscribers vs Customers in the data frame. ***
```{r count-usertypes, echo=FALSE}
table(divvy_data$usertype)
```
***I need to load the csv file named "Divvy_Trips_2019_Q2.csv" and 2019_Q3 2019_Q4 into R and print the first 6 rows of the data frame.***
```{r load-data-data}
divvy_data_q2 <- read.csv("Divvy_Trips_2019_Q2.csv", header = TRUE)
divvy_data_q3 <- read.csv("Divvy_Trips_2019_Q3.csv", header = TRUE)
divvy_data_q4 <- read.csv("Divvy_Trips_2019_Q4.csv", header = TRUE)
head(divvy_data_q2)
head(divvy_data_q3)
head(divvy_data_q4)
```
*** adding a collumn for quarter to all four datasets indicating the quarter ***
```{r add-quarter-column}
divvy_data$quarter <- "Q1"
divvy_data_q2$quarter <- "Q2"
divvy_data_q3$quarter <- "Q3"
divvy_data_q4$quarter <- "Q4"
```

*** Combine all four quarters of Divvy trip data into a single data frame. ***
```{r combine-trip-data}
divvy_data_all <- rbind(divvy_data, divvy_data_q2, divvy_data_q3, divvy_data_q4)
head(divvy_data_all)
```

*** Check the structure difference between the four dataframes ***
```{r check-structure}
names(divvy_data)
names(divvy_data_q2)
names(divvy_data_q3)
names(divvy_data_q4)  
```